---
title: "Hybrid Model for the Age of Infection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hybrid Model for the Age of Infection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, suppressMessages=T}
library(deSolve)
library(pf.memory)
library(knitr)
library(viridisLite)
library(viridis)
```

```{r, echo=F}
a2years = c(0:730)
foiP3 = list(hbar = 5/365, agePar = par_type2Age(), seasonPar = par_sinSeason(), trendPar = par_flatTrend()) 
clrs = turbo(12)
```


In [AoI](AoI.html), we discuss the direct computation of the moments of the distribution of the AoI: 

+ Let $x$ denote the first moment of the AoI in a cohort at age $a$, conditioned on a hestory of exposure: $A_\tau(a; h_\tau(a))$: $$x_\tau(a; h) = \left< A_\tau(a; h) \right> = \int_0^\infty \alpha \frac{z_\tau(\alpha, a; h_\tau(a))} {m_\tau(a;h_\tau(a))}$$

+ Similarly, we let $x_\tau(a)[n]$ denote the higher order moments of $A_\tau(a;)$: $$x_{n}(a, \tau; h) = \int_0^\infty \alpha^n \frac{z_\tau(\alpha, a; h_\tau(a))} {m_\tau(a;h_\tau(a))}$$

Here, we derive hybrid variables describing changes in the first moments: 

$$
\frac{\textstyle{dx}}{\textstyle{da}} = 1- 
\frac{\textstyle{h}}{\textstyle{m}} x  
$$

All subsequent moments can be computed recursively: 

$$
\frac{\textstyle{dx_n}}{\textstyle{da}} = nx_{n-1} -  
  \frac{\textstyle{h}}{\textstyle{m}} x_n 
$$

In the following, we first demonstrate that the hybrid equations give the same answers as direct computation of the moments. 

## Demo

```{r, purl=F, fig.height=6, fig.width=7}
par(mar = c(7, 4, 2, 2))
solve_dAoI(5/365, foiP3) -> mt 
plot(mt$time, mt$xn1, type = "l", xlab = "Age (in Days)", ylab = expression(E(A[tau](a))), ylim = range(mt$xn3^(1/3))) 
lines(mt$time, mt$xn2^(1/2))
lines(mt$time, mt$xn3^(1/3)) 
```

```{r}
aa = seq(1, 5*365, by = 5) 
moment1 = momentAoI(aa, foiP3)
moment2 = momentAoI(aa, foiP3, n=2)
moment3 = momentAoI(aa, foiP3, n=3)
```

2. By solving a hybrid model with variables that track the moments, using the equations that we derived above: 

```{r} 
solve_dAoI(5/365, foiP3, Tmax = 5*365, dt=5) -> mt 
```

```{r, fig.height=10, fig.width=6}
par(mfrow = c(3,1), mar = c(7, 4, 2, 2))
plot(mt$time, mt$xn1, type = "l", xlab = "Age (in Days)", ylab = expression(E(A[tau](a))), lwd=2) 
lines(aa, moment1, col = "red", lwd=2, lty=2)
plot(mt$time, mt$xn2, type = "l", xlab = "Age (in Days)", ylab = "second moment", lwd=2) 
lines(aa, moment2, col = "red", lwd=2, lty=2)
plot(mt$time, mt$xn3, type = "l", xlab = "Age (in Days)", ylab = "third moment", lwd=2)
lines(aa, moment3, col = "red", lwd=2, lty=2)
```

## Derivation for $dx/da$

In [Dynamics](Dynamics.html), we present an equation for   $z_\tau(\alpha,a),$ the density function for infections of age $\alpha$ in a host cohort of age $a$ born on day $\tau$. The dynamics of $z(\alpha,a)$ are described by the following:

\begin{equation}
\tag{1}
\frac{\partial z}{\partial a} + \frac{\partial z}{\partial \alpha} = - r z
\end{equation}

with the boundary condition:

$$
    z_\tau(0,a) = h_\tau(a)
$$

and its solutions are: 

$$
z_\tau(\alpha,a) = h_\tau(a-\alpha) e^{-r \alpha}
$$

In [MoI](MoI.html), we defined the distribution function for the MoI: 

$$M_\tau(a) \sim f_M(\zeta; a, \tau) = \mbox{Pois}(m_\tau(a))$$

and we showed that the mean MoI, $m_\tau(a)$ is:   

\begin{equation}
m_\tau(a) = \int_0^a z_\tau(\alpha, a) d \alpha
\end{equation}

In [AoI](AoI.html), we defined a probability density function for the AoI: 

$$
A_\tau(a) \sim f_A(\alpha; a, \tau) = \frac{z_\tau(\alpha,a)}{m_\tau(a)}
$$

Here, we derive an equation for $$\frac{dx}{da} = \frac{d\left<A\right>}{da}$$ 

We start by multiplying Eq. 1 by $\alpha$ and integrating with respect to $\alpha$: 

\begin{equation}
\int_0^a \alpha \frac{\partial z(\alpha, a)}{\partial a} d \alpha + \int_0^a \alpha \frac{\partial z(\alpha, a)}{\partial \alpha} d\alpha = - r \int_0^a \alpha z(\alpha, a) d\alpha, 
\end{equation}

By its definition, 

$$\int_0^a \alpha z(\alpha, a) d\alpha =  \left<A \right> m$$

So we can simplify a bit:  

\begin{equation}
\tag{2}
\frac{\partial \left<A \right>m}{\partial a} d \alpha + \int_0^a \alpha \partial z  = - r \left<A \right>  m 
\end{equation}

On the RHS of Eq. 2, the first term can be rewritten: 

\begin{equation}
\frac{\partial \left<A \right> m}{\partial a}= 
m \frac{\partial \left<A \right>}{\partial a}  + 
 \left<A \right> \frac{\partial m}{\partial a}  
\end{equation}

On the RHS of Eq. 2, the second term can be rewritten by integrating by parts: 

\begin{equation}
\int_0^a \alpha \partial z = 
\alpha z(a,\alpha) \big|_{\alpha = a} - 
\alpha z(a,\alpha) \big|_{\alpha = 0} - 
\int_0^a z d\alpha 
\end{equation}

which simplifies to:

\begin{equation}
\int_0^a \alpha \partial z =  a h_d(0)e^{-ra} - m
\end{equation}

Substituting, we get: 

\begin{equation}
m \frac{\partial \left<A \right>}{\partial a}  +  \left<A \right> \frac{\partial m}{\partial a}  
 + a h_d(0)e^{-ra} - m  = - r \left<A \right>  m 
\end{equation}



We divide through by $m$ and rearrange to get: 

\begin{equation}
 \frac{\partial \left<A \right>}{\partial a}  =  - \frac{\left< A \right>}{m} \frac{\partial m}{\partial a}
- \frac{a}{m} h_d(0)e^{-ra} + 1  - r \left<A \right>   
\end{equation}

We substitute $dm/da = h - rm$ and simplify: 

\begin{equation}
 \frac{\partial \left<A \right>}{\partial a}  =  - \frac{\left< A \right>}{m} (h - rm) 
- \frac{a}{m} h_d(0)e^{-ra} + 1  - r \left<A \right>   
\end{equation}

and after rearranging terms and cancelling:

\begin{equation}
 \frac{\partial \left<A \right>}{\partial a}  =  1 - h \frac{\left< A \right>}{m} 
- \frac{a}{m} h_d(0)e^{-ra} 
\end{equation}

or since we are letting $x = \left< A \right>$ 
\begin{equation}
 \frac{d x}{d a}  =  1 -  \frac{h}{m} x 
- \frac{a}{m} h_d(0)e^{-ra}   
\end{equation}

*I can't remember why that last term goes away*

Noting that:
\begin{equation}
    \lim_{m\rightarrow 0} \frac{A}{m} =  \lim_{m\rightarrow 0} \frac{a}{m} = 1 
\end{equation}

## Derivation for $dx_n/da$


To derive a tracking equation for $\left<A^n \right>$, we start as before but multiply by $\alpha^n$ to get:

\begin{equation}
\int_0^a \alpha^n \frac{\partial z}{\partial a} \partial \alpha + \int_0^a \alpha^n \frac{\partial z}{\partial \alpha} \partial \alpha = \int_0^a - r \alpha^n z \partial \alpha     
\end{equation} 
which we can rewrite as: 
\begin{equation}
\frac{\partial \left<A^n \right> m}{\partial a} + \int_0^a \alpha^n \partial x  = - r \left<A^n\right> m 
\end{equation} 

Integrating the middle term by parts,   

\begin{equation}
\int_0^a \alpha^n \partial z = 
\alpha^n z(a,\alpha) \big|_{\alpha = 0}^{a} - 
n \int_0^a \alpha^{n-1} z d\alpha 
\end{equation} 

so we can rearrange this to get Eq.~\ref{dAnda}. Once again, we are concerned with the limits as $m$ becomes small: 

\begin{equation}
    \lim_{m\rightarrow 0} \frac{A}{m} =  \lim_{m\rightarrow 0} \frac{a}{m} = 1 
\end{equation}
