---
title: "Parasite Density Distributions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parasite Density Distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r suppressMessages=T}
library(pf.memory)
library(viridisLite)
```

In the following, we build the functions to compute parasite density distributions using information about the MoI and the aoI. To do so:

1. $F_\mu$ -- We developed the `S3` function `alpha2mu` to build models of the relationship between mean $\log_{10}$ parasite densities ($\mu$) and the AoI ($\alpha$); $$\mu = F_\mu(\alpha)$$ 

2. $\Omega$ -- We develop four `S3` functions to compute the density, distribution, and quantile functions and random numbers for parasite density distributions, given mean $\log_{10}$ densities, $\mu$: `d_mu2density`, `p_mu2density`, `q_mu2density` and `r_mu2density` $$\Omega(\xi; \mu)$$ 

3. $P$ - We develop functions to compute the density distribution of parasites in simple infections in a human cohort of age $a$: 
$$
P_\tau(a | h) \sim  f_P(\xi; a, \tau |h ) = \int_0^a \Omega(\xi|\mu(\alpha)) \; f_A(\alpha; a, \tau | h) d\alpha 
$$
4. $B$ - We develop functions to compute the density distributions of parasites in a human cohort of age $a$  
$$
B_\tau(a|h) \sim f_B(\xi; a, \tau |h) = \log_{10} \left( \sum_{M_\tau(a|h)>0} 10^{P_\tau(a |h)}\right)
$$
The software is modular, in the sense that $F_\mu$ and $\Omega$ can use any suitable function, and $h$ is passed as a trace function.

## Expected Logged Densities 

In `pf.memory,` a `S3` function family was developed to model the relationship between the age of infection (AoI) for a parasite infection, $\alpha$, and the mean, $\mu(\alpha).$ 

### `alpha2mu.base` 

The function `alpha2mu.base` has the following form:  

\begin{equation}
F_\mu(\alpha) = 
\begin{cases}
    \text{NA} \ & \mbox{if } 0 \leq \alpha < 7 \\[10pt]
      l + (b - l)  \frac{{\textstyle \alpha}}{{\delta}}   & \mbox{if } 7 \leq \alpha \leq \delta  \\[10pt]
     l + (b - l) e^{-s_\alpha (\alpha-\delta)}   & \mbox{if } \alpha \geq \delta\\  
\end{cases}
\label{mualpha}
\end{equation}

### The function `alpha2mu.chronic` has the following form:  

\begin{equation}
F_\mu(\alpha) = 
l + (b - l) e^{-s_\alpha (\alpha-\delta)}   
\end{equation}


In `pf.memory,` there are functions to set up the parameters for each case, with over-writable defaults: 

```{r}
base_pars = par_Fmu_base()
base_pars1 = par_Fmu_base(tildeb=11, Sa=0.0045)
chronic_pars = par_Fmu_chronic()
```


```{r}
aalpha = 7:366
mu_alpha =  alpha2mu(aalpha, 0, base_pars)
mu_alpha1 =  alpha2mu(aalpha, 0, base_pars1)
mu_alpha2 =  alpha2mu(aalpha, 0, chronic_pars)
```

We plot the two functions here: with the default parameters (in black), and the other parameters (in dark blue):

```{r, fig.height=4, fig.width=6, echo=F}
plot(aalpha, mu_alpha, type = "l", ylim = c(0, 13), xlab = expression(alpha), ylab = expression (mu(alpha)))
lines(aalpha, mu_alpha1, col = "darkblue")
lines(aalpha, mu_alpha2, col = "darkred")
segments(0,6,365,6, lty=2)
```
## Parasite Density Distributions 

We have developed a function family to compute parasite densities as a function of the mean of logged parasite densities, $\mu$


```{r}
parD = par_mu2dens_beta() 
d_mu2density(9, 9, parD)
```

```{r, fig.height=8, fig.width=8}
par(mfrow = c(2,2))
xx = seq(0, 13, by=0.1)
plot(xx, d_mu2density(xx, 9, parD), type = "l", 
     xlab = expression(xi), ylab = "d_mu2density", main = "d_mu2density")
lines(xx, d_mu2density(xx, 7,  parD), col = "darkred")
pp = seq(0,13, by = 0.01)
plot(pp, p_mu2density(pp, 9,  parD), type = "l", 
     xlab = expression(xi), ylab = "p_mu2density", main = "p_mu2density")
lines(xx, p_mu2density(xx, 7, parD), col = "darkred")
hist(r_mu2density(1000, 9, parD), xlab = expression(xi), 
     main = "r_mu2density", xlim = c(0,13))
qq = seq(0.01,.99, length.out=100)
plot(qq, q_mu2density(qq, 9,  parD), type = "l", 
     xlab = expression(xi), ylab = "q_mu2density", main = "q_mu2density")
```

## Simple Infections 

We first look at the density of parasites in simple infections -- those with MoI = 1. 

To compute anything, we need a function to model exposure, the FoI as a function of age, $h_\tau(a)$.  Notably, the function accepts the birth date, $\tau$ as an argument, so it can be used to look at cohorts with different birth dates. Here, we show the FoI as a function of age for two cohorts born six months apart. 

```{r, fig.height=4, fig.width=7, echo=F}
foiP3 = list(hbar = 5/365, 
             agePar = par_type2Age(), 
             seasonPar = par_sinSeason(), 
             trendPar = par_flatTrend())

a5years  = 0:(5*365)
clrs = turbo(2)
plot(a5years, FoI(a5years, foiP3, tau=0), type = "l", col = clrs[1],
     xlab = "a - host cohort age (in days)", ylab = expression(FoI[tau](a)))
lines(a5years, FoI(a5years, foiP3, tau=180), col = clrs[2])
#lines(a5years, FoI(a5years, foiP3, tau=270), col = clrs[4])
```


For one of these cohorts (the black line), we have plotted the density parasite cohorts. Notably, the age of the cohorts are representing infections that occurred $\alpha$ days ago, so the $x-$axis is looking into the past:  

```{r}
mu_alpha =  alpha2mu(a5years, 0, base_pars)
dAoI(a5years, 5*365, foiP3) -> dalpha
```

```{r, echo=F, fig.height=4, fig.width=7, echo=F}
plot(a5years, dalpha, type = "l", 
    xlab = expression(list(alpha, paste("Parasite Age (in Days)"))), 
    ylab = expression(f[A](alpha)))
```

Now we want to understand parasite densities in the cohort of humans. To compute the full probability distribution function in that cohort of 5-year olds, we use both bits of information:

+ Mean logged parasite densities for a parasite infection by age, $\mu(\alpha)$ 

+ The density of cohorts by age, $A_\tau(a)$

```{r, echo=F, fig.height=4.5, fig.width=7, echo=F}
par(mar = c(5,5,3,4))
dd = dalpha*max(mu_alpha)/max(dalpha)
plot(a5years, mu_alpha, type = "l", ylim = c(0,11),
    xlab = expression(list(alpha, paste("Parasite Age (in Days)"))), 
    ylab = expression(list(xi, paste("Parasite Densities"))), 
    main = expression(list(mu(alpha), f[A]())))

lines(a5years, dd, col = grey(0.8), lwd=2)

clrs = rev(turbo(50))
i = 1:50
points(20*i, dd[20*i], col =clrs, pch = 15, cex=0.7) 
tks = pretty(dalpha)
axis(4, tks*max(mu_alpha)/max(dalpha), tks)
mtext(expression(f[A](mu)), 4,3)
```

For a parasite of any age, we can plot its probability density function. Here, we show the probability density function for a parasite infection at age $60$ days:

```{r}
alpha = 60
fA = dAoI(alpha, 5*365, foiP3)
Pmu = fA*d_alpha2density(xx, alpha, 5*365)
```

```{r, echo=F, fig.height=4, fig.width=7, echo=F}
plot(xx, Pmu, type = "l", ylim = range(Pmu)*1.2)
```



Now, parasite densities in simple infections have the given distribution of mean logged parasite densities. To compute the full distribution, we create a mesh over $\xi$:

**For every age of infection, $\alpha$**

+ We compute expected parasite densities $\mu(\alpha)$

+ We compute the full parasite density distribution, multiplied by the density of parasites of age $\alpha$: $$\phi(\alpha) = P(\mu(\alpha))\; f_A(\alpha, a)$$

+ We add $\phi(\alpha)$ to its appropriate bin, defined by the mesh over $\xi$. 

The function `dPdensity` computes the parasite distributions as a function of the host age, given a function describing the history of exposure.


```{r}
Pa = d_Pdensity(xx, 5*365, foiP3)
```

```{r, echo=F, fig.height=4.5, fig.width=7}

parD = par_mu2dens_beta()
alpha = 20
mu = alpha2mu(alpha, 0, base_pars)
fA = dAoI(alpha, 5*365, foiP3)
Pmu = fA*d_mu2density(xx, mu, parD)

par(mar = c(5,5,3,4))
plot(xx, Pa, type = "l",
     ylab = expression(P[a](xi)),
     xlab = expression(xi),
     main = "Parasite Density Distributions" )
segments(6,0, 6, 0.004, col = grey(0.5), lty=2)
for(i in (1:50)){
  alpha = 1020-20*i
  mu = alpha2mu(alpha, 0, base_pars)
  fA = dAoI(alpha, 5*365, foiP3)
  lines(xx, fA*d_mu2density(xx, mu, parD)*max(Pa)/max(Pmu), col = clrs[51-i])
}
tks = pretty(Pa/40, 6)
axis(4, tks*40, tks)
mtext(expression(P[mu](alpha)), 4, 3)
```

We can check to ensure that our distribution function integrates to 1: 

```{r}
integrate(d_Pdensity, 0, 13, a=5*365, FoIpar=foiP3)$value 
```

## Complex Infections




