---
title: "Prevalence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prevalence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(pf.memory)
```

```{r}
library(knitr)
library(deSolve)
library(pf.memory)
```

George Maccdonald published a model of superinfection in 1950^[G Macdonald (1950). The analysis of infection rates in diseases in which superinfection occurs. Trop. Dis. Bull. 47, 907–915.].
In presenting the model, Macdonald was trying to solve a problem: if super-infection was possible, then how long would it take until a person would clear all of the parasites to become uninfected again? 
Macdonald proposed a solution, but there was a problem with the mathematical formulation (An essay by Paul Fine, from 1975, is recommended reading)^[Fine PEM (1975). Superinfection - a problem in formulating a problem. Tropical Diseases Bulletin 75, 475–488]:
the model that Macdonald described didn't match the function he used. 
In a model formulated as part of the Garki project (Dietz K, *et al.*,  1974)^[Dietz K, Molineaux L, Thomas A (1974). A malaria model tested in the African savannah. Bull. Wld. Hlth. Org. 50, 347–357.], 
the model allowed for *superinfection* and proposed a useful approximating model. 
One solution came from taking a hybrid modeling approach (Nåsell I, 1985)^[Nåsell I (1985). Hybrid Models of Tropical Infections, 1st edition. Springer-Verlag. https://doi.org/10.1007/978-3-662-01609-1], which we have discussed in a vignette about the [Multiplicity of Infection (MoI)](MoI.html). 

The model Macdonald described was also formulted in queuing theory, where the model is called $M/M/\infty$. In this model, each new infection increases the MoI by one; the transition rate to a higher MoI is the force of infection (FoI), denoted $h$. Each parasite can clear at some rate, $r$. 

The following diagram shows how the fraction of the population with MoI $=i$, denoted $\zeta_i$, changes over time.  

***

\begin{equation}
\begin{array}{ccccccccc}
\zeta_0 &  {h\atop \longrightarrow} \atop {\longleftarrow \atop r} & \zeta_1 
& {h\atop \longrightarrow} \atop {\longleftarrow \atop {2r}} & \zeta_2 
& {h \atop \longrightarrow} \atop {\longleftarrow \atop {3r}} & \zeta_3 
& {h \atop \longrightarrow} \atop {\longleftarrow \atop {4r}}& \ldots
\end{array}
\end{equation}

***

While the infinite system of differential equations can be formulated and solved numerically, Nåsell showed that the system can be described by a simple equation. Let $m$ denote the mean MoI. 

$$
\frac{dm}{dt} = h-rm
$$

He also showed that the distribution of the MoI would converge to a Poisson distribution: 

$$M(t) \sim f_M(\zeta; m) = \mbox{Pois}(m(t))$$


